Visual Question Answering with DeepProbLog

### 2022-03-24 paper/paper.tex
Final review of paper

\section{INTRODUCTION}
The Neuro Symbolic AI field is interested in building a bridge between the robustness of probabilistic knowledge, with the well-known popularity and proven strengths of deep neural networks. DeepProbLog \cite{deepproblog} offers this ability, by using both the strengths of neural networks (i.e. system 1, typical subconscious tasks such as visual recognition, the processing of languages, \dots), along with the strengths of rule-based probabilistic systems (i.e. system 2, slow, sequential thinking such as the derivation of a proof).

@@

\section{ORGANIZATION OF THE PAPER}
This paper will first provide some necessary background in Section \ref{sec:literature_survey} on certain types of datasets that are typically used for Visual Question Answering purposes, along with introducing some results of other researchers. Then, Section \ref{sec:approach} will dive deeper in the different components constituting the overall system. A main focus of this paper is to outline the advantages of using a Neuro-Symbolic AI approach (offered by the DeepProbLog framework), compared to a purely neural network based approach. These experiments are listed in Section \ref{sec:experiments}. Finally, the main findings of this paper are repeated in Section \ref{sec:conclusions}.

@@

This paper is inspired on the CLEVR dataset, but uses however a much more simplified version. In essence, it is almost like the Sort-Of-CLEVR dataset \cite{sort_of_clevr_dataset}. This Sort-Of-CLEVR dataset contains images as in Figure \ref{fig:sample_image_sort_of_clevr}, while asking questions such as:
\begin{itemize}
    \item Non-relational questions: the shape, horizontal or vertical location of an object.
    \item Relational questions: shape of the closest/furthest object to the object under investigation, or the number of objects with the same shape.
\end{itemize}

\begin{figure}[htp]
    \begin{center}
    \fbox{\includegraphics[width=0.25\textwidth]{sort_of_clevr.png}}
    \captionsetup{justification=centering}
    \caption{A sample image from the CLEVR dataset \cite{sort_of_clevr_dataset}}
    \label{fig:sample_image_sort_of_clevr}
    \end{center}
\end{figure}

@@

\section{APPROACH}
\label{sec:approach}
The implementation process involved three main parts:
\begin{enumerate}
    \item Generation of data.
    \item Linking the data and controlling the training process in pure Python code.
    \item Creation of the logical part with DeepProbLog statements.
\end{enumerate}

\subsection{Generation of data}
As mentioned in Section \ref{sec:literature_survey}, the data used in this application is based on the Sort-Of-CLEVR dataset, with one extra simplification. Given that the logical part will have to decide whether an object is for example located on the left side of an image, the neural network will have to convey positional information to the logical part. Hence, each discrete position will have to be encoded by a possible outcome of the neural network. Therefore, objects may only be located at certain positions in a grid. In this paper, results on a grid of 2x2 and 6x6 are discussed.

The data generator that was used for the creation of the Sort-Of-CLEVR dataset has been modified in order to place objects in the mentioned grid positions \cite{sort_of_clevr_dataset}. An example of a generated image is given in Figure \ref{fig:sample_image_own_dataset}, where the difference with Figure \ref{fig:sample_image_sort_of_clevr} is the grid-layout.

@@

These images are accompanied with a question about a random object, which can be one of the following:
\begin{itemize}
    \item Non-relational - What is the shape of this object?\footnote{The shape will be either a square or a circle.}
    \item Non-relational - Is this object located on the left side of the image?
    \item Non-relational - Is this object located on the bottom side of the image?
    \item Relational - How many objects have the same shape as this object?
\end{itemize}

@@

\section{EXPERIMENTS}
\label{sec:experiments}
The main focus of this paper is to outline the advantages of using a Neuro Symbolic AI approach (offered by the DeepProbLog framework), instead of a purely neural network based approach. Therefore, a neural network based approach had to be implemented. Not all details will be listed here, but the main idea is that the image has to be fused with the question, after which a prediction can be made\footnote{Possible answers on the questions are: rectangle, circle, yes, no, 1, 2, 3, 4, 5 and 6. Hence, the neural network will have 10 output nodes.}. The general structure of this network is given in Figure \ref{fig:pure_nn_network}.

@@

\subsection{COMPARISONS WITH PURE SYSTEM 1 APPROACHES}
\label{subsec:experiments_NN_vs_deepproblog}
\subsubsection{Experiment - 2x2 grid}
The loss curves of both the DeepProbLog approach, as well as the purely neural network based approach, are visualized respectively in Figure \ref{fig:loss_curve_deepproblog_2x2} and Figure \ref{fig:loss_curve_pure_NN_2x2}.

@@

\subsubsection{Experiment - 6x6 grid}
The loss curves for the 6x6 experiment of DeepProbLog and the neural network based approach are depicted in respectively Figure \ref{fig:loss_curve_deepproblog_6x6} and Figure \ref{fig:loss_curve_pure_NN_6x6}.

@@

In these experiments, the training time of the purely neural network approach was 20 minutes\footnote{For fair speed estimation, training has been conducted on the CPU.}, while the training time of the DeepProbLog approach was a little under eight hours\footnote{However, it should be mentioned that approximately half of the time was spent on calculating the accuracy each time, since the whole testing set had to be forwarded through the network, while for the training iterations this is only one batch.}.

The most important observation here is that the purely neural network approach overfits quickly and also achieves a considerably lower accuracy (68\% instead of almost 75\% for DeepProbLog). Another important remark is the fact that both approaches are clearly not longer able to attain 100\% accuracy. However, if the DeepProbLog network could train longer, it would be able to converge somewhat further.

@@

Another observation is that DeepProbLog is much better in answering the questions `What shape has the object under investigation?' and `Is the object located on the left hand side (or on the bottom side) of the image?' then a purely neural network approach is. This makes sense, since DeepProbLog can use its knowledge base to derive these properties, once the position (and shape) of the given object is determined. The observation that a pure neural network based approach has a much harder time to distinguish these cases, was also observed by A. Santoro et al. \cite{sort_of_clevr_dataset}, where they achieved an accuracy of 52.3\% on these questions with a pure neural network based approach. DeepProbLog was not able to achieve the accuracy of 94\% that these researchers achieved, which may be due to the inherent training cost of the algebraic operators, as well as due to less resources and less hyperparameter tuning, among many other unknown variables.

@@

\section{CONCLUSIONS}
\label{sec:conclusions}
The strengths of the Neuro Symbolic AI field have been demonstrated in the context of Visual Question Answering. By using DeepProbLog, the chosen framework for the Neuro Symbolic AI task, it became clear that almost 200 times less iterations are required to achieve the same accuracy as a purely neural network based approach. However, due to the costly algebraic operators, the total training time of the DeepProbLog approach was considerably slower, compared to the purely neural network based approach.

@@

Hence, a lot of value can be seen in Neuro Symbolic AI approaches, despite the costly algebraic operators. Especially because for tasks where the knowledge base is much larger, these approaches can make the difference between being able to learn a certain task or not. Over time, speedups for these algebraic operators could probably be developed, which opens the road to even more applications.

>>>>

\section{INTRODUCTION}
The Neuro Symbolic AI field is interested in building a bridge between the robustness of probabilistic knowledge, with the well-known popularity and proven strengths of deep neural networks. DeepProbLog \cite{deepproblog} offers this ability, by using both the strengths of neural networks (i.e. system 1, typical subconscious tasks such as visual recognition, the processing of languages, \dots), along with the strengths of rule-based probabilistic systems (i.e. system 2, slow, sequential thinking such as the derivation of a proof) \cite{sys1and2}.

@@

\section{ORGANIZATION OF THE PAPER}
This paper will first provide some necessary background in Section \ref{sec:literature_survey} on certain types of datasets that are typically used for Visual Question Answering purposes, along with introducing some results of other researchers. Then, Section \ref{sec:approach} will dive deeper in the different components constituting the overall system. A main focus of this paper is to outline the advantages of using a Neuro Symbolic AI approach (offered by the DeepProbLog framework), compared to a purely neural network based approach. These experiments are listed in Section \ref{sec:experiments}. Finally, the main findings of this paper are repeated in Section \ref{sec:conclusions}.

@@

This paper is inspired on the CLEVR dataset, but uses however a much more simplified version. In essence, it is almost like the Sort-of-CLEVR dataset \cite{sort_of_clevr_dataset}. This Sort-of-CLEVR dataset contains images as in Figure \ref{fig:sample_image_sort_of_clevr}, while asking questions such as:
\begin{itemize}
    \item Non-relational questions: the shape, horizontal or vertical location of an object.
    \item Relational questions: shape of the closest/furthest object to the object under investigation, or the number of objects with the same shape.
\end{itemize}

\begin{figure}[htp]
    \begin{center}
    \fbox{\includegraphics[width=0.25\textwidth]{sort_of_clevr.png}}
    \captionsetup{justification=centering}
    \caption{A sample image from the Sort-of-CLEVR dataset \cite{sort_of_clevr_dataset}}
    \label{fig:sample_image_sort_of_clevr}
    \end{center}
\end{figure}

@@

\section{APPROACH}
\label{sec:approach}
The implementation process involved three main parts:
\begin{enumerate}
    \item Generation of the data.
    \item Linking the data and controlling the training process in pure Python code.
    \item Creation of the logical part with DeepProbLog statements.
\end{enumerate}

\subsection{Generation of the data}
As mentioned in Section \ref{sec:literature_survey}, the data used in this application is based on the Sort-of-CLEVR dataset, with one extra simplification. Given that the logical part will have to decide whether an object is for example located on the left side of an image, the neural network will have to convey positional information to the logical part. Hence, each discrete position will have to be encoded by a possible outcome of the neural network. Therefore, objects may only be located at certain positions in a grid. In this paper, results on a grid of 2x2 and 6x6 are discussed.

The data generator that was used for the creation of the Sort-of-CLEVR dataset has been modified in order to position objects in the mentioned grid positions \cite{sort_of_clevr_dataset}. An example of a generated image is given in Figure \ref{fig:sample_image_own_dataset}, where the difference with Figure \ref{fig:sample_image_sort_of_clevr} is the grid-layout.

@@

These images are accompanied with a question about a random object, which can be one of the following:
\begin{itemize}
    \item Non-relational --- What is the shape of this object\footnote{The shape will be either a square or a circle.}?
    \item Non-relational --- Is this object located on the left hand side of the image?
    \item Non-relational --- Is this object located on the bottom side of the image?
    \item Relational --- How many objects have the same shape as this object?
\end{itemize}

@@

\section{EXPERIMENTS}
\label{sec:experiments}
The main focus of this paper is to outline the advantages of using a Neuro Symbolic AI approach (offered by the DeepProbLog framework), instead of a purely neural network based approach. Therefore, a neural network based approach had to be implemented. Not all details will be listed here, but the main idea is that the image has to be fused with the question, after which a prediction can be made\footnote{Possible answers on the questions are: rectangle, circle, yes, no, 1, 2, 3, 4, 5 and 6. Hence, the neural network will have 10 output nodes. Care has been taken for class skew, by introducing weights in the loss function.}. The general structure of this network is given in Figure \ref{fig:pure_nn_network}.

@@

\subsection{COMPARISONS WITH PURE SYSTEM 1 APPROACHES}
\label{subsec:experiments_NN_vs_deepproblog}
\subsubsection{Experiment --- 2x2 grid}
The loss curves of both the DeepProbLog approach, as well as the purely neural network based approach, are visualized respectively in Figure \ref{fig:loss_curve_deepproblog_2x2} and Figure \ref{fig:loss_curve_pure_NN_2x2}.

@@

\subsubsection{Experiment --- 6x6 grid}
The loss curves for the 6x6 experiment of DeepProbLog and the neural network based approach are depicted in respectively Figure \ref{fig:loss_curve_deepproblog_6x6} and Figure \ref{fig:loss_curve_pure_NN_6x6}.

@@

In these experiments, the training time of the purely neural network based approach was 20 minutes\footnote{For fair speed estimation, training has been conducted on the CPU.}, while the training time of the DeepProbLog approach was a little under eight hours\footnote{However, it should be mentioned that approximately half of the time was spent on calculating the accuracy each time, since the whole testing set had to be forwarded through the network, while for the training iterations this is only one batch.}.

The most important observation here is that the purely neural network approach overfits quickly and also achieves a considerably lower accuracy (68\% instead of 75\% for DeepProbLog). Another important remark is the fact that both approaches are clearly not longer able to attain an accuracy of 100\%. However, if the DeepProbLog network could train longer, it would be able to converge somewhat further.

@@

Another observation is that DeepProbLog is much better in answering the questions `What shape has the object under investigation?' and `Is the object located on the left hand side (or on the bottom side) of the image?' than a purely neural network approach is. This makes sense, since DeepProbLog can use its knowledge base to derive these properties, once the position (and shape) of the given object is determined. The observation that a pure neural network based approach has a much harder time to distinguish these cases, was also observed by A. Santoro et al. \cite{sort_of_clevr_dataset}, where they achieved an accuracy of 63\% on these questions with a pure neural network based approach. DeepProbLog was not able to achieve the accuracy of 94\% that these researchers achieved on all questions with an added RN module, which may be due to the inherent training cost of the algebraic operators, as well as due to less resources, less hyperparameter tuning, and an absence of RN modules, among many other unknown variables.

@@

\section{CONCLUSIONS}
\label{sec:conclusions}
The strengths of the Neuro Symbolic AI field has been demonstrated in the context of Visual Question Answering. By using DeepProbLog, the chosen framework for the Neuro Symbolic AI task, it became clear that almost 200 times less iterations are required to achieve the same accuracy as a purely neural network based approach. However, due to the costly algebraic operators, the total training time of the DeepProbLog approach was considerably slower, compared to the purely neural network based approach.

@@

Hence, a lot of value can be seen in Neuro Symbolic AI approaches, despite the costly algebraic operators. Especially for tasks where the knowledge base is much larger, these approaches can make the difference between being able to learn a certain task or not. Over time, speedups for these algebraic operators could probably be developed, which opens the road to even more applications.
